Process-based phenotypes: an application to the Cellular Phenotype Ontology

### 2011-11-23 paper/drinking.tex

The second common type of abnormality are abnormalities based on
process participants in relation to the duration of the process. These
include all kinds of {\em rates} such as {\em mass flow rate}, {\em
  energy flow rate} and {\em velocity} (the rate of change of
position). In each of these cases, an object participates in a process
and a quality (or change of quality) of that object throughout the
duration of the process is considered to form a new quality. If the
process has participants that are distinguished into {\em inputs} and
{\em outputs}, then a recurring pattern is that the amount of inputs
or outputs that participate in the process can be {\em increased} or
{\em decreased}. For example, an {\em Increased rate of cytoplasmic
  streaming} is defined as...

@@

A second kind of regulatory abnormalities is related to the {\em
  onset} of a process. With respect to a reference process, a
particular kind of process may be {\em delayed} ({\tt PATO:0000502})
or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
  apoptosis} refers to an abnormality of the {\em Regulation of
  apoptosis} in which apoptosis is induced later than normal.  We use
the PATO quality {\em onset} ({\tt PATO:...}) and its children {\em
  delayed} and {\em premature} to define these types of regulatory
abnormality. Similarly, we use PATO's {\em offset ({\tt PATO:...})
  quality and its children to characterize regulatory abnormalities in
  which a process ends prematurely or too late.

\begin{table*}
  \begin{tabular}{l|l|l|l}
    & Increased blood flow rate & Normal blood flow rate & Decreased
    blood flow rate \\
    Increased frequency &increased&increased&?\\
    Normal frequency &increased&normal&decreased\\
    Decreased frequency &?&decreased&decreased
  \end{tabular}
  \caption{\label{...}Total blood flow rate...}
\end{table*}

Finally, a third kind of regulatory abnormality refers to abnormal
rates with respect to a participant of the process that is being
regulated. For example, a cytoplasmic flow rate can

>>>>

The second common type of abnormality are abnormalities based on
process participants in relation to the duration of the process. These
include all kinds of {\em rates} such as {\em mass flow rate}, {\em
  energy flow rate} and {\em velocity} (the rate of change of
position). In each of these cases, an object participates in a process
and a quality (or change of quality) of that object throughout the
duration of the process is considered to form a new quality. If the
process has participants that are distinguished into {\em inputs} and
{\em outputs}, then a recurring pattern is that the amount of inputs
or outputs that participate in the process can be {\em increased} or
{\em decreased}. For example, an {\em Increased rate of cytoplasmic
  streaming} can be defined as an increased amount of inputs or an
increased amount of outputs of a {\em cytoplasmic streaming} process.

@@

A second kind of regulatory abnormalities is related to the {\em
  onset} of a process. With respect to a reference process, a
particular kind of process may be {\em delayed} ({\tt PATO:0000502})
or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
  apoptosis} refers to an abnormality of the {\em Regulation of
  apoptosis} in which apoptosis is induced later than normal.  We use
the PATO quality {\em onset} ({\tt PATO:...}) and its children {\em
  delayed} and {\em premature} to define these types of regulatory
abnormality. Similarly, we use PATO's {\em offset} ({\tt PATO:...})
  quality and its children to characterize regulatory abnormalities in
  which a process ends prematurely or too late.

\begin{table*}
  \centering
  \begin{tabular}{l|l|l|l}
    & Increased cytoplasmic flow rate & Normal cytoplasmic flow rate & Decreased
    cytoplasmic flow rate \\
    \hline
    Increased frequency &increased total flow rate &increased total
    flow rate &?\\
    Normal frequency &increased total flow rate &normal total flow
    rate &decreased total flow rate\\
    Decreased frequency &?&decreased total flow rate &decreased total
    flow rate\\
    \hline
  \end{tabular}
  \caption{\label{tbl:flow}Total cytoplasmic flow rate...}
\end{table*}

Finally, a third kind of regulatory abnormality refers to abnormal
rates with respect to a participant of the process that is being
regulated. For example, a cytoplasmic flow rate can be increased or
decreased not within a single {\em cytoplasmic streaming} process but
rather the total cytoplasmic flow rate, as a summation over all
cytoplasmic streaming processes that occur within an organism (or a
particular anatomical location), is increased or decreased. While a
flow rate of a single {\em cytoplasmic streaming} process is a quality
of that process, an increased {\em total} cytoplasmic flow rate is a
quality of the regulation of {\em cytoplasmic streaming}. In
particular, it is possible for an organism to have a normal --- or
even a decreased --- cytoplasmic flow rate in each individual
cytoplasmic streaming process while at the same time having an
increased total cytoplasmic flow rate due to a large increase in the
frequency of occurrence of cytoplasmic streaming processes. Similarly,
the frequency of occurrence of cytoplasmic streaming may be normal or
decreased while the total cytoplasmic flow rate is increased due to an
increased cytoplasmic flow rate in each individual cytoplasmic
streaming process. Table \ref{tbl:flow} illustrates the dependencies
between rates of individual processes, their frequency of occurrence
and the total rate of these processes. We include total rates as
regulatory abnormalities in the CPO since these are the attributes of
processes that are often measured or observed, while the rates of
individual processes are inferred following a schema such as Table
\ref{tbl:flow}.
